---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.retina = TRUE
)

library(dplyr)
library(phewasHelper)

# Simulate some random data associated with phecodes
phewas_data <- sample_n(phecode_descriptions, 300) %>%
  left_join(mutate(category_colors(return_df = TRUE),
                   bias = rnorm(n(), sd = 3)),
            by = 'category') %>% 
  mutate(val = rnorm(n(), mean = bias),
         code = as.numeric(phecode)) %>% 
  select(code, val)
```

# phewasHelper

<!-- badges: start -->
<!-- badges: end -->

The goal of phewasHelper is to provide a set of simple lightweight helper functions for working with Phecode and Phewas data. 

## Installation

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nstrayer/phewas_helper")
```



## Usage

We will use a sample set of simulated PheWas data with phecodes to demonstrate the functions. 


```{r show-data}
library(phewasHelper)

head(phewas_data)
```



### Normalizing phecodes

Phecodes show up in about a million different formats. In our demo data the phecodes have been converted to a numeric value. This would be an issue if we tried to harmonize with data that stored the phecodes in a string. The function `normalize_phecodes` is designed to fix this problem. It takes any phecode array and coerces it to a standard zero-padded string. 

```{r normalize_phecodes}
phewas_data %>% 
  mutate(fixed_code = normalize_phecode(code)) %>% 
  head()


# Update our original data with normalized phecodes
phewas_data <- phewas_data %>% 
  mutate(code = normalize_phecode(code))
```

### Getting phecode information

Another issue that is commonly encountered in PheWas results is wanting to know what exactly a code is. The functions `get_phecode_info()` and `join_phecode_info()` help with that. 

`get_phecode_info()` is the simpler of the two. It takes as input an array of phecodes and returns an array of the desired information, either description or category. This is useful for adding individual columns to a dataframe. 

```{r get_phecode_info}
phewas_data %>% 
  mutate(descript = get_phecode_info(code, 'description'),
         category = get_phecode_info(code, 'category')) %>% 
  head()
```

For more a more complete labeling of phecode information the function `join_phecode_info()` modifies a passsed dataframe by appending a desired subset of description, category, category number, and phecode index columns. 

```{r join_phecode_info}
# We can append all info available
phewas_data %>% 
  join_phecode_info(phecode_column = 'code') %>% 
  head()


# Or we can just extract what we need
phewas_data <- phewas_data %>% 
  join_phecode_info(phecode_column = 'code',
                    cols_to_join = c("description", "category", "phecode_index"))

head(phewas_data)
```


### Coloring PheWas plots

Manhattan plots are commonly made of phewas results. Frequently the colors of the plots points are encoded by the categories. The default color palletes in ggplot2 and base-plot are not great and custom palletes like R color-brewer don't give you enough colors to work with all the categories. To deal with this `category_colors()` returns a mapping of phecode category to colors that can be used easily in your plots. 

```{r plot-w-category_colors}
library(ggplot2)

phewas_data %>% 
  ggplot(aes(x = reorder(phecode, phecode_index), y = val, color = category)) +
  geom_point() +
  scale_color_manual(values = category_colors()) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```


If just the color pallete is needed for `ggplot` then the function `scale_color_phecode()` makes this even easier. 

```{r plot-w-scale_color_phecode}
phewas_data %>% 
  ggplot(aes(x = reorder(phecode, phecode_index), y = val, color = category)) +
  geom_point() +
  scale_color_phecode() +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

